{\small} High Performance P$^{3}$M N-body code: CUBEP3M

### 2012-06-21 Extendedpp.tex
Updated force plots with rebins and N=10 averages

 To speed up the access to all particles within the domain of computation, we construct a thread safe linked list
 to be constructed and accessed in parallel by each core of the system, but this time with a head-of-chain that points to the first particle in the current fine mesh cell. We then loop over all fine grids, accessing the particles contained therein and inside each fine grid cells for which we killed the mesh kernels,
 we compute the separation and the force between each pairs and update their velocities simultaneously with Newton's third law. 
 To avoid double counting, we loop only over the fine mesh neighbours that produce non-redundant contributions. Namely, for a central cell located at 
 $(x_1, y_1, z_1)$, we only consider the neighbours $(x_2, y_2, z_2)$ that satisfy the following conditions:
 \begin{itemize}
 \item{$z_2 \ge z_1$ always}
 \item{if $z_2 = z_1$, then $y_2 \ge y_1$, otherwise we also allow $y_2 < y_1$} 
 \item{if $z_2 = z_1$ and $y_2 = y_1$, then we enforce $x_2 > x_1$}
 \end{itemize}
 The case where all three coordinates are equal is already calculated in the standard configuration of the code.
 To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_fracErr_ppext6} a force versus distance
 plot in a CITA256 realization , analogous to Fig. \ref{fig:den_force_fracErr}, but the pp force has been extended to six layers of fine cells. 
 We observe that the scatter about the theoretical curve has reduced significantly, and is still well balanced around the theoretical predictions.
 The same applies to the fractional error on the radial and tangential components of the force, as seen in the right panel,
 which are now at least 5 time smaller than in the default P$^{3}$M algorithm.

 \begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=6.eps}
%    \label{fig:den_force_ppext6}}
%\end{center}
%\end{figure}
%\begin{figure}%[ht]
%  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6.eps}
    \caption{({\it left}:) Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
  in a CITA256 realization. The exact pp force has been extended to six fine mesh layers around each particle.
  Particles in the that range follow the exact curve, then we observe a much smaller scatter at 
  distance of the order of 6 fine grid, compared to Fig. \ref{fig:den_force_fracErr}. 
  The NGP interpolation scheme is again responsible for the scatter, but the effect is suppressed with increasing distances, and further suppressed by averaging over many time steps.
  %\caption{
  ({\it right}:) Fractional error on the force in the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions,
  calculated from a CITA256 simulation in which the exact pp force extends to six fine mesh layers. 
  This was also obtained over a single time step.
    \label{fig:den_force_fracErr_ppext6}}
\end{center}
\end{figure*}

>>>>

 To speed up the access to all particles within the domain of computation, we construct a thread safe linked list
 to be constructed and accessed in parallel by each core of the system, but this time with a head-of-chain that points to the first particle in the current fine mesh cell. We then loop over all fine grids, accessing the particles contained therein and inside each fine grid cells for which we killed the mesh kernels,
 we compute the separation and the force between each pairs and update their velocities simultaneously with Newton's third law. 
 To avoid double counting, we loop only over the fine mesh neighbours that produce non-redundant contributions. Namely, for a central cell located at 
 $(x_1, y_1, z_1)$, we only consider the neighbours $(x_2, y_2, z_2)$ that satisfy the following conditions:
 \begin{itemize}
 \item{$z_2 \ge z_1$ always}
 \item{if $z_2 = z_1$, then $y_2 \ge y_1$, otherwise we also allow $y_2 < y_1$} 
 \item{if $z_2 = z_1$ and $y_2 = y_1$, then we enforce $x_2 > x_1$}
 \end{itemize}
 The case where all three coordinates are equal is already calculated in the standard configuration of the code.
 To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_fracErr_ppext6} a force versus distance
 plot in a CITA128 realization, analogous to Fig. \ref{fig:den_force_fracErr}, but the pp force has been extended to  six layers of fine cells
 in the force test part of the code {\bf (do pp\_ext2 as well?)}. 
 We observe that the scatter about the theoretical curve has reduced significantly, and is still well balanced around the theoretical predictions.
 The same applies to the fractional error on the radial and tangential components of the force, as seen in the right panel,
 which are now at least 5 time smaller than in the default P$^{3}$M algorithm {\bf (Check these numbers again)}.
 When averaging over 10 time steps, the improvement is only mild, as the calculation is already very accurate.

 \begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=6_rebin.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6_rebin.eps}
     \includegraphics[width=3.2in]{graphs/densityForce_ppext=6_rebin_N10.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6_rebin_N10.eps}
    \caption{({\it top left}:) Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
  in the same CITA128 realization as that shown in Fig. \ref{fig:den_force_fracErr}, 
  except that the exact pp force has been extended to six fine mesh layers around each particle
  in the force test code.
  Particles in the that range follow the exact curve, then we observe a much smaller scatter at 
  distance of the order of 6 fine grid, compared to Fig. \ref{fig:den_force_fracErr}. 
  The NGP interpolation scheme is again responsible for the scatter, but the effect is suppressed with increasing distances, and further suppressed by averaging over many time steps.
  %\caption{
  ({\it top right}:) Fractional error on the force in the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
  This was also obtained over a single time step.
  ({\it bottom row}:) Same as top row, but averaged over ten time steps.
{ \bf (I should add pp\_ext = 2. ) }
    \label{fig:den_force_fracErr_ppext6}}
\end{center}
\end{figure*}

### 2012-06-21 Halos.tex
Updated force plots with rebins and N=10 averages

We have implemented a halo finding procedure, which we have developed 
based on the spherical overdensity (SO) approach \citep{1994MNRAS.271..676L}.
In the interest of speed and efficiency the halo catalogues are constructed 
on-the-fly at a pre-determined list of redshifts. The halo finding is 
massively-parallel and threaded based on the main {\small CUBEP3M} data structures 
discussed in section \ref{sec:structure}. The code first builds the 
fine-mesh density for each sub-domain using CIC or NGP interpolation. It then 
proceeds to search for and record all local density maxima above certain
threshold (typically set to 100 above mean density) within the local 
sub-domain's physical volume (excluding the sub-domain buffer zones). It then 
uses parabolic interpolation on the density field to determine more precisely
the location of the maximum within the densest cell, and records the peak 
position and value. The halo centre determined this way agrees closely with 
the centre-of-mass of the halo particles discussed below.

Once the list of peak positions is generated, they are sorted from the highest 
to the lowest density value. Then each of the halo candidates is inspected 
independently, starting with the highest peak. The grid mass is accumulated 
in spherical shells of fine grid cells surrounding the maximum, until the 
mean density within the halo drops below a pre-defined overdensity cutoff 
(usually set to 178 in units of the mean, in accordance to the top-hat 
collapse model). As we accumulate mass we remove it from the mesh, so that no 
mass element is double-counted. This method is thus inappropriate for finding 
sub-halos as within this framework those are naturally incorporated in their 
host halos. Because the halos are found on a grid of finite-size cells, it is 
possible, especially for the low-mass halos, to overshoot the target overdensity.
When this occurs we use an analytical halo density profile to correct the 
halo mass and radius to the values corresponding to the target overdensity. 
This analytical density profile is given by Truncated Isothermal Sphere (TIS) 
profile \citep{1999MNRAS.307..203S,2001MNRAS.325..468I} for overdensities below 
$\sim130$, and $1/r^2$ for lower overdensities. The density profile has a
similar outer slope (the relevant one here) to the Navarro, Frenk and White 
\citep[NFW][]{1997ApJ...490..493N} profile, but extends to lower overdensities
and matches well the virialization shock position given by the Bertschinger 
self-similar collapse solution \citep{1985ApJS...58...39B}.

Once the correct halo mass, radius and position are determined, we find all 
particles which are within the halo radius. Their positions and velocities are
used to calculate the halo centre-of-mass, bulk velocity, internal velocity 
dispersion and the three angular momentum components, all of which are then 
included in the final halo catalogues. We also calculate the total mass of
all particles within the halo radius, also listed in the halo data. This mass
is very close, but  typically slightly lower, than the halo mass calculated 
based on the gridded density field. The particle centre-of-mass corresponds 
very well to the halo centre found based on the gridded mass distribution.

@@

\begin{figure}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.2in]{graphs/mf_z0_Tinker.eps}
  \end{center}
  \caption{Simulated halo multiplicity function, 
    $\frac{M^2}{\bar{\rho}}\frac{dn}{dM}$ based on a
    RANGER4000 simulation with $3.2\,h^{-1} \mbox{Gpc}$ box and $4000^3$ 
    particles (solid, red in the online version). For reference we also show a widely-used 
    precise fit by \citet{2008ApJ...688..709T} (blue, dashed). 
    \label{mf}}
\end{figure}

A sample halo mass function produced based on our inlined SO halo finder 
at redshift $z=0$ from a RANGER4000 simulation is shown in Fig. \ref{mf}. We compare our result to the 
precise fit presented recently by \citet{2008ApJ...688..709T}. Unlike most
other widely-used fits like the one by \citet{2002MNRAS.329...61S}, which are based on friends-of-friends (FOF)
halo finders, the \citet{2008ApJ...688..709T} fit is based on the
SO search algorithm, whose masses are systematically different 
from the FOF masses \citep[e.g.][]{2007MNRAS.374....2R,2008ApJ...688..709T}, 
making this fit a better base for comparison here. Results show excellent
agreement, within $\sim10$ per cent for all halos with masses corresponding to
1000 particles or more. Lower-mass halos are somewhat under-counted compared
to the \citet{2008ApJ...688..709T} fit, by $\sim20$ per cent for 400 particles and 
by $\sim40$ per cent for 50 particles. This is due to the grid-based nature of our
SO halo finder, which misses some of the low-mass halos. It was shown, however, that using more sophisticated
halo finders (available through post-processing due to their heavier memory
footprint) allows to recover the expected mass function.

>>>>

We have implemented a halo finding procedure, which we have developed 
based on the spherical overdensity (SO) approach \citep{1994MNRAS.271..676L}.
In the interest of speed and efficiency the halo catalogues are constructed 
on-the-fly at a pre-determined list of redshifts. The halo finding is 
massively-parallel and threaded based on the main {\small CUBEP3M} data structures 
discussed in section \ref{sec:structure}. The code first builds the 
fine-mesh density for each sub-domain using CIC or NGP interpolation. It then 
proceeds to search for all local density maxima above a certain
threshold (typically set to a factor of 100 above mean density) within the local tile. 
It then uses parabolic interpolation on the density field to determine more precisely
the location of the maximum within the densest cell, and records the peak 
position and value. %The halo centre determined this way agrees closely with 
%the centre-of-mass of the halo particles discussed below.

Once the list of peak positions is generated, they are sorted from the highest 
to the lowest density value. Then each of the halo candidates is inspected 
independently, starting with the highest peak. The grid mass is accumulated 
in spherical shells of fine grid cells surrounding the maximum, until the 
mean density within the halo drops below a pre-defined overdensity cutoff 
(usually set to 178 in units of the mean, in accordance with the top-hat 
collapse model). As we accumulate the mass, we remove it from the mesh, so that no 
mass element is double-counted. This method is thus inappropriate for finding 
sub-haloes since within this framework, those are naturally incorporated in their 
host haloes. Because the haloes are found on a discrete grid, it is 
possible, especially for those with lower mass, to overshoot the target overdensity.
To minimize this efect, we correct the halo mass and radius with an analytical density profile. 
We use the Truncated Isothermal Sphere (TIS) 
profile \citep{1999MNRAS.307..203S,2001MNRAS.325..468I} for overdensities below 
$\sim130$, and a simple $1/r^2$ law for lower overdensities {\bf (lower or higher?)}. 
These yield a similar outer slope to the Navarro, Frenk and White 
\citep[NFW][]{1997ApJ...490..493N} profile, but extends to lower overdensities
and matches well the virialization shock position given by the Bertschinger 
self-similar collapse solution \citep{1985ApJS...58...39B}.

Once the correct halo mass, radius and position are determined, we find all 
particles that are within the halo radius. Their positions and velocities are
used to calculate the halo centre-of-mass, bulk velocity, internal velocity 
dispersion and the three angular momentum components, all of which are then 
included in the final halo catalogues. We also calculate the total mass of
all particles within the halo radius, also listed in the halo data. This mass
is very close, but  typically slightly lower, than the halo mass calculated 
based on the gridded density field. The  centre-of-mass found this way closely follows 
that found from the peak location, which is based on the gridded mass distribution.

@@

\begin{figure}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.2in]{graphs/mf_z0_Tinker.eps}
  \end{center}
  \caption{Simulated halo multiplicity function, 
    $\frac{M^2}{\bar{\rho}}\frac{dn}{dM}$, based on a
    RANGER4000 simulation with $3.2\,h^{-1} \mbox{Gpc}$ box and $4000^3$ 
    particles (solid, red in the online version). For reference we also show a widely-used 
    precise fit by \citet{2008ApJ...688..709T} (blue, dashed). 
    \label{mf}}
\end{figure}

A sample halo mass function produced  from a RANGER4000 simulation at $z=0$ is shown in Fig. \ref{mf}. We compare our result to the 
precise fit presented recently by \citet{2008ApJ...688..709T}. Unlike most
other widely-used fits like the one by \citet{2002MNRAS.329...61S}, which are based on friends-of-friends (FOF)
halo finders, this one is based on the
SO search algorithm, whose masses are systematically different 
from the FOF masses \citep[e.g.][]{2007MNRAS.374....2R,2008ApJ...688..709T}, 
making this fit a better base for comparison here. Results show excellent
agreement, within $\sim10$ per cent for all haloes with masses corresponding to
1000 particles or more. Lower-mass haloes are somewhat under-counted compared
to the \citet{2008ApJ...688..709T} fit, by $\sim20$ per cent for 400 particles and 
by $\sim40$ per cent for 50 particles. This is due to the grid-based nature of our
SO halo finder, which misses some of the low-mass haloes. It was shown, however, that using more sophisticated
halo finders (available only through post-processing calculations due to their heavier memory
footprint) it is possible to recover the expected mass function.

### 2012-06-21 ScalingPerformance.tex
Updated force plots with rebins and N=10 averages

Furthermore, due to the increasing clustering of structures at those small 
scales, some of the cuboid sub-domains came to contain a number of particles
well above the 
average, thereby requiring more memory per {\small MPI} process
in order to run until the end. 
As a consequence, {\it throughout most of the evolution}{\bf (Ilian, did you actually restarted simulations and switched 4ways to 2ways?)} the 
largest two of these simulations were run with 4096 and 21,952 cores and 
with only 2 {\small MPI} processes and 8 threads per Ranger node (`2way'), which on 
Ranger allows using up to 16 GB of RAM per {\small MPI} process\footnote{In order to insure 
local memory affinity,  a second special NUMA control script, {\it tacc\_affinity\_2way}) 
was developed for us by the TACC technical staff and allowed to run more efficiently 
in this mode.}. Because each processor accesses memory that is not fully local, this configuration does affect the 
performance somewhat, as does the imperfect load balancing that arises in such situations.
This can be seen in the rightmost point of FIg. \ref{scaling} (right panel), where the scaling 42 per cent below the ideal.
We note that we still get $\sim1.5$ speedup
from doubling the core count, even given these issues. Overall the
code scaling performance is thus satisfactory even at extremely large number of cores.
 We expect the code to handle even larger problems efficiently, and is thus well suited to run on
  next generation Petascale systems.

>>>>

Furthermore, due to the increasing clustering of structures at those small 
scales, some of the cuboid sub-domains came to contain a number of particles
well above the 
average, thereby requiring more memory per {\small MPI} process
in order to run until the end. 
As a consequence,  throughout most of their late evolution, the 
largest two of these simulations were run with 4096 and 21,952 cores and 
with only 2 {\small MPI} processes and 8 threads per Ranger node (`2way'), which on 
Ranger allows using up to 16 GB of RAM per {\small MPI} process\footnote{In order to insure 
local memory affinity,  a second special NUMA control script, {\it tacc\_affinity\_2way}, 
was developed for us by the TACC technical staff and allowed to run more efficiently 
in this mode.}. Because each processor accesses memory that is not fully local, this configuration does affect the 
performance somewhat, as does the imperfect load balancing that arises in such situations.
This can be seen in the rightmost point of FIg. \ref{scaling} (right panel), where the scaling 42 per cent below the ideal.
We note that we still get $\sim1.5$ speedup
from doubling the core count, even given these issues. Overall the
code scaling performance is thus satisfactory even at extremely large number of cores.
 We expect the code to handle even larger problems efficiently, and is thus well suited to run on
  next generation Petascale systems.

### 2012-06-21 Systematics.tex
Updated force plots with rebins and N=10 averages

\section{Accuracy}
\label{sec:accuracy}

One of the most reliable way to assess the simulation's accuracy at evolving particles
is to measure the density power spectrum at late time, and compare to non-linear prediction. 
For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
\begin{eqnarray}
\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
\label{eq:power}
\end{eqnarray}
where the angle bracket corresponds to an ensemble (or volume) average.
We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum in a RANGER4000 simulation and
observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ are caused by the white noise imposed in the initial conditions. {\b (I should add a vertical line at the k of the coarse mesh for visualization purposes.)}

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=5.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation,
  compared to the non-linear predictions of {\small CAMB}.
    \label{fig:power_highres}}
\end{center}
\end{figure*}

The actual force of gravity in the P$^3$M algorithm,
as felt by a single particle, is presented in Fig. \ref{fig:den_force_ppext0}.
This shows the force versus distance -- in fine cell units -- a calculation that was performed from a CITA256 realization in two steps: 
1- we compute the force on each particle in a given time step.
2- we remove a selected particle, compute the force again on all particles, and record on file the 
force difference (before  and after the removal) as a function of the distance to the `hole'.

Particles in the same fine cell as the hole follow the exact $1/r^{2}$ curve. The scatter at 
  distances of the order of the fine grid is caused by the NGP interpolation scheme:
  particles in adjacent fine cells can be actually very close, as seen in the upper left region of this plot,
  but still feel the mesh force at grid cell distances,
  creating up to an order of magnitude loss.
As the separation approaches a tenth of the full box size or so, the force on the coarse mesh deviates
from Newton's law in order to preserve periodic boundary conditions. 
This feature, however, can be turned off at compilation time.

Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial and tangential directions.
It is larger than in {\small PMFAST} at grid scales, largely due to the fact that the fine mesh force is performed with an NGP interpolation scheme -- as opposed to CIC -- which shows a larger scatter about the theoretical value. 
As discuss in the next section, to minimize this effect, we apply a random offset to all the particles,
such that this undesired scatter is averaged out over a few time steps.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
  \caption{Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
    This particular calculation was obtained in a CITA256 realization with  a box size of $500 h^{-1}\mbox{Mpc}$,
    in a single time step. In a full simulation run, the scatter averages over many time steps, thanks to the inclusion of a random offset that is imposed on each particle. 
    \label{fig:den_force_ppext0}}
\end{center}
\end{figure}

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
  \caption{Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
  This scatter is significantly reduced with the random offset, as discussed in section \ref{sec:systematics}.
    \label{fig:den_force_fracErr}}
\end{center}
\end{figure}

\section{Systematics}
\label{sec:systematics}

\subsection{Mesh force at grid distances}

The biggest problem with the straightforward pp force calculation is that the results 
are anisotropic and depend on the location of the fine mesh with respect 
to the particles. As an example, consider two particles on either side of a grid 
cell boundary, experiencing their mutual gravity attraction via the fine mesh force with a discretized 1-grid cell separation.
 If, instead, the mesh was shifted such that they were
within the same cell, they would experience the much larger pp force. 
This effect is especially pronounced at the early stages of the simulation where
the density is more homogeneous, and leads to mesh artefacts appearing
in the density field. In order to minimize this systematic effect, 
we randomly shift the particle distribution relative to the mesh by a small
amount -- up to 2 fine grid cells in magnitude -- in each
dimension and at each time step.  This adds negligible computational
overhead as it is applied during the particle position update,
and suppresses the mesh behaviour that otherwise grows over multiple time steps.
It is possible to shift back the particles at the end of each time steps,
which prevents a random drift of the whole population, a necessary step 
if one needs to correlate the initial and final positions of the particles for instance,
or for hybrid dark matter -- MHD simulations.

We ensure that, on average, this solution balances out the mesh feature,
by tuning the force kernels such as to provide  force as evenly balanced as possible at grid cell distances
and at the cutoff length ($r_{c}=16$ fine cells).
These adjustments are performed from the pairwise force test described in section \ref{sec:accuracy}.
We note that this is one of the driving argument to extend the pp force outside the fine mesh cell,
since the scattering of the NGP force about the actual $1/r^{2}$ law drops rapidly as the distance increases.
As discussed in section \ref{subsec:extendedpp}, this gain in accuracy comes at a price,
a choice that  must be carefully balanced.

We present in Fig. \ref{fig:disp_mesh} the dramatic impact of removing the random offset in the code.
The power spectrum is completely wrong, due to the large scatter in the force from the fine mesh.
In this scatter is not averaged over, the error directly adds up at each time step. {\small PMFAST}
did not have this problem since it used CIC interpolation on both meshes.  
This test was performed with CITA256 simulations of very large box size,
which should in principle agree with linear predictions up to the resolution.
The output redshifts are very early, and the upper most curve ($z=10$) was obtained after
about 60 time steps.
We see that deviations are enormous when the random offset is turned off.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_w_wo_disp_mesh.eps}
  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a series of CITA256 realizations that are 1000 Mpc/$h$ per side. The dashed line represent the initial condition power spectrum, the dotted lines are the linear predictions, and  the open circles the standard P$^3$M configuration. 
  The dots were obtained by simply removing the random particle offset that is usually applied at each time step. \label{fig:disp_mesh}}
\end{center}
\end{figure}

{\bf (Show how averaging over $~sim10$ time steps reduces the scatter.)}

>>>>

\section{Accuracy and Systematics}
\label{sec:accuracy}

 This section describes the systematics effects that are inherent to our P$^{3}$M algorithm.
 We start with a demonstration of the accuracy with a power spectrum measurement of a RANGER4000 simulation. 
 The halo mass function also assess the code capacity to model gravitational collapse, 
 but depends on the particular halo finder used.
 We thus post-pone the discussion on this aspect until in section \ref{sec:halo},
and focus for the moment on the particles only. In addition to the power spectrum, we further quantify the accuracy of the force calculation
 with comparisons to Newton's law of gravitation.

 \subsection{Power Spectrum}
 \label{subsec:powerspectrum}

One of the most reliable way to assess the simulation's accuracy at evolving particles
is to measure the density power spectrum at late time, and compare to non-linear prediction. 
For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
\begin{eqnarray}
\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
\label{eq:power}
\end{eqnarray}
where the angle bracket corresponds to an ensemble (or volume) average.
We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum in a RANGER4000 simulation and
observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
The drop of power at high-$k$ is partly caused by the finite mesh resolution, partly from expected deviations about the predictions.
The fluctuations at low-$k$ come from the white noise imposed in the initial conditions. The vertical line corresponds to the scale of the coarse mesh.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=5.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation,
  compared to the non-linear predictions of {\small CAMB}. 
  The vertical line corresponds to the scale of the coarse mesh, while the dotted line represents the Poisson noise.
    \label{fig:power_highres} {\bf Should I show another redshift instead? Also, the legend box cuts the text: need to fix this. Finally, I could also show linear theory.}}
\end{center}
\end{figure*}

\subsection{Mesh force at grid distances}
\label{subsec:force}

The force test presented in MPT was carried by placing two particles at random locations on the grid, calculate the force between them, then iterate
over different locations. This pairwise force test is useful to quantify the accuracy on a cell-by-cell basis, but lacks the statistics that occur in an actual time step calculation.  
The actual force of gravity in the P$^3$M algorithm,
as felt by a single particle during a single step, is presented in the top left panel of Fig. \ref{fig:den_force_fracErr}.
This force versus distance plot was obtained from a CITA128 realization, and the calculation proceeds in two steps: 
1- we compute the force on each particle in a given time step.
2- we remove a selected particle, thus creating a `hole', compute the force again on all particles, and record on file the 
force difference (before  and after the removal) as a function of the distance to the hole.

Particles in the same fine cell as the hole follow the exact $1/r^{2}$ curve. The scatter at 
  distances of the order of the fine grid is caused by the NGP interpolation scheme:
  particles in adjacent fine cells can be actually very close, as seen in the upper left region of this plot,
  but still feel the same mesh force at grid cell distances.
This  creates a discrepancy up to an order of magnitude, in loss or gain, depending on the location of the pair with respect to the centre of the cell.
As the separation approaches a tenth of the full box size or so, the force on the coarse mesh deviates
from Newton's law in order to preserve periodic boundary conditions.

It is not clear from this scatter plot whether the mean force felt  is below or under the predictions, hence we rebin the results
in 50 logarithmically spaced bins and compute the mean and standard deviation. 
We show this plot in the middle left panel of Fig. \ref{fig:den_force_fracErr}, where we observe that on average,
there is a loss of force of the order of unity in the range $ 0.4 < r < 1$ (in units of fine grid), but the force force is otherwise
in excellent agreement with Newton's law.

The top right and middle right panels of Fig. \ref{fig:den_force_fracErr} show the fractional error on the force along the radial and tangential directions, also calculated from a single time step.
The mean of the radial component is up to 80 per cent away from the exact law at sub-grid distances, but quickly drops under ten per cent for larger distances. 
The transverse fractional error peaks exactly at the gris distance,  and is about two time smaller than the radial counterpart.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_rebin.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_rebin.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_rebin_N10.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_rebin_N10.eps}
  \caption{({\it top left}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
    This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$,
    in a single time step. 
    ({\it top right}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
  In a full simulation run, the scatter averages over many time steps, 
    thanks to the inclusion of a random offset that is imposed on each particle, as discussed in the main text.
    ({\it middle row}:) Top row organized in 50 logarithmic bins.  
    ({\it bottom row}:) Same as middle row, but averaged over ten time steps.
    \label{fig:den_force_fracErr}
    {\bf (Why is frac err radial slightly under 0? Problem in the units?)}}
\end{center}
\end{figure*}

At grid scales, the fractional error is larger than {\small PMFAST}, largely due to the fact that the fine mesh force is performed with an NGP interpolation scheme -- as opposed to CIC. This prescription is responsible for the larger scatter about the theoretical value, but, as mentioned earlier, 
NGP interpolation is essential to our implementation of the pp part.
At the same time, the biggest problem with the straightforward pp force calculation is that the results 
are anisotropic and depend on the location of the fine mesh with respect 
to the particles. As an example, consider two particles on either side of a grid 
cell boundary, experiencing their mutual gravity attraction via the fine mesh force with a discretized one-grid cell separation.
 If, instead, the mesh was shifted such that they were
within the same cell, they would experience the much larger pp force. 
This is clearly seen in the top left panel of Fig. \ref{fig:den_force_fracErr}, where particles physically close, but in different grid cells, 
feel a force up to an order of magnitude too small.
This effect is especially pronounced at the early stages of the simulation where
the density is more homogeneous, and leads to mesh artefacts appearing
in the density field {\bf (Should I show a density plot?)}.

In order to minimize these two systematic effect -- the large scatter and the anisotropy -- 
we randomly shift the particle distribution relative to the mesh by a small
amount -- up to 2 fine grid cells in magnitude -- in each
dimension and at each time step.  This adds negligible computational
overhead as it is applied during the particle position update,
and suppresses the mesh behaviour that otherwise grows over multiple time steps.
It is possible to shift back the particles at the end of each time steps,
which prevents a random drift of the whole population, a necessary step 
if one needs to correlate the initial and final positions of the particles for instance,
or for hybrid dark matter -- MHD simulations.

We ensure that, on average, this solution balances out the mesh feature,
by tuning the force kernels such as to provide a force as evenly balanced as possible, both at grid cell distances
and at the cutoff length ($r_{c}=16$ fine cells).
These adjustments are performed from the pairwise force test mentioned above (and described in MPT).
The bottom panels of Fig. \ref{fig:den_force_fracErr} show the effect of averaging over ten time steps, 
where the random offset is applied on the particles (and on the hole) at the end of each force calculation. 
We observe that the agreement extends at the per cent level down to $r \sim 1.5$, beyond which point
the discretize effect of the NGP causes an underestimate of the force of up to 60 per cent.
{\bf (Can't we beat this by tempering with the fine force kernel? Will the least square fit enhance this?)}

We note that this inevitable loss of force in the NGP scheme is one of the driving argument to extend the pp force outside the fine mesh cell,
since the scattering about the actual $1/r^{2}$ law drops rapidly as the distance increases.
As discussed in section \ref{subsec:extendedpp}, this gain in accuracy comes at a computational price,
but at least we have the option to run the code with high precision.

We present in Fig. \ref{fig:disp_mesh} the dramatic impact of removing the random offset in the otherwise default code configuration.
This test was performed with a CITA256 simulation of very large box size,
the output redshifts are very early (the upper most curve ($z=10$) was obtained after
about 60 time steps), such that the power spectrum should agree with linear theory up to the resolution limit.
Instead, we observe that the power spectrum grows completely wrong, due to the large scatter in the force from the fine mesh,
and to the anisotropic nature of the P$^3$M calculations mentioned above.
When these effects are not averaged over, the errors directly add up at each time step,
which explains why later time are worst.
We recall that {\small PMFAST} did not have this problem since it used CIC interpolation on both meshes.  
{\bf (Anything else to add here?)}

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_w_wo_disp_mesh.eps}
  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a CITA256 realization that is 1000 Mpc/$h$ per side. The dashed line represent the initial condition power spectrum, the dotted lines are the linear predictions, and  the open circles the standard P$^3$M configuration. 
  The dots were obtained by simply removing the random offset that is normally applied at each time step. \label{fig:disp_mesh}}
\end{center}
\end{figure}

%\begin{figure*}%[ht]
 % \begin{center}
  %   \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N1.eps}
   % \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N10.eps}
  % \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N1.eps}
  %  \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N10.eps}
 % \caption{({\it top}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
 %   This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$.
  %  ({\it bottom}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
   % The left panels are calculated from a single time step, while the right panels show the average over time time steps.
 %   \label{fig:den_force_N10}}
%\end{center}
%\end{figure*}

%\begin{figure}%[ht]
 % \begin{center}
  %   \includegraphics[width=3.2in]{graphs/densityForce_ppext=10_rebin.eps}
  %\caption{ Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
   % This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$, averaged over 10 time steps.
   % Results are rebined to show how well the mean value follows the thoretical curve down to the grid size.
   % \label{fig:F_rebin}}
%\end{center}
%\end{figure}
